Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-09-25 ms.tex
copyediting

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time. Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions. For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while --- to ensure reproducibility --- past versions of the dataset would remain accessible.

Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

@@

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system. The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}). Our focus here is on the wide range of datasets not covered by these repositories, such as those appearing in this journal. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not in most cases general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Likewise, while the examples cited here come from our own research areas the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.

@@

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses. A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.  
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

Currently there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.  This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

@@

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solution. But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions and involve standard data types (e.g., sequences). But for an individual research group with expertise on a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

>>>>

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will change through time. Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions. For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.

However, current models for publishing datasets do not facilitate distributing successive versions of a dataset in an effective and scalable manner. The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}). Once published, the datasets are immutable. That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published. While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset. Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.

@@

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system. The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}). Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not in most cases general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Likewise, while the examples cited here come from our own research areas, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.

@@

To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be transparent and understandable by both data sharers and users.  Crucially a user should be immediately aware of the relationship among versions.  Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.  Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution. This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions. Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results. A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses. A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns. These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.  
% I think that major version increments could still be looked at as
% _breaking_ changes as often as large additions of data.  Changing
% column names, removing columns, changing the structure of linked
% tables etc.

@@

Currently there are two types of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.  The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).  \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).  Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.  This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.

@@

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solutions. But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.
